Penalized regression for left-truncated and right-censored survival data

### 2021-04-19 manuscript.tex
minor edits- Efron approx, HR bias

where $\beta$ is the vector of regression coefficients and is maximized over \autoref{eqn:simon} subject to a constraint defined by $\lambda \cdot P_{\alpha}\beta$, and $P_{\alpha}\beta$ is the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty on the sum of coefficients $\beta$ described elsewhere \cite{simon2011regularization, tibshirani1996regression, tikhonov1963ridge, zou2005regularization}. The algorithm by Simon \emph{et al.} fits over a path of $\lambda$ values via cyclical coordinate descent. Instead of solving a least squares problem at each iteration, the algorithm solves a penalized reweighted least squares problem, in which observation weights $w$ are defined in the coordinate descent least squares minimization step. The optimal $\lambda$ can be identified through cross-validation. Tied survival times are handled via the Breslow approximation \cite{breslow1972} and details provided in \cite{simon2011regularization}.

@@

In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor "year of diagnosis" moved from being the largest in absolute value to around $1$ after adjusting for left truncation. The strongly positive effect of "year of diagnosis" on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. This effect is reversed completely after the left truncation adjustment, and illustrates the bias to coefficient estimates that is driven by the association between predictors and left truncation time. In the case of "year of diagnosis," this association is negative and drives an upward bias to the hazard ratio. While many of the other hazard ratios in the large $p$ model were pushed towards $1$ after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor. Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.

>>>>

where $\beta$ is the vector of regression coefficients and is maximized over \autoref{eqn:simon} subject to a constraint defined by $\lambda \cdot P_{\alpha}\beta$, and $P_{\alpha}\beta$ is the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty on the sum of coefficients $\beta$ described elsewhere \cite{simon2011regularization, tibshirani1996regression, tikhonov1963ridge, zou2005regularization}. The algorithm by Simon \emph{et al.} fits over a path of $\lambda$ values via cyclical coordinate descent. Instead of solving a least squares problem at each iteration, the algorithm solves a penalized reweighted least squares problem, in which observation weights $w$ are defined in the coordinate descent least squares minimization step. The optimal $\lambda$ can be identified through cross-validation. Tied survival times are handled via the Breslow approximation \cite{breslow1972} and details provided in \cite{simon2011regularization}. Note that the computationally efficient Efron approximation will be added to the next version of \pkg{glmnet}.

@@

In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor "year of diagnosis" moved from being the largest in absolute value to around $1$ after adjusting for left truncation. The strongly positive effect of "year of diagnosis" on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. This effect is reversed completely after the left truncation adjustment, and illustrates that biases to coefficient estimates are driven by the association between predictors and left truncation time. In the case of "year of diagnosis," this association is negative and biases the hazard ratio upwards. While many of the other hazard ratios in the large $p$ model were pushed towards $1$ after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor. In cases where predictors and left truncation time are positively associated, hazard ratios are biased downwards because the predictors are associated with more immortal time.
